---
title: "Dynamic Fertility Modeling Exercises"
subtitle: "Berkeley Formal Demography Workshop"
author: "Joshua Goldstein, Ayesha Mahmud, Paige Park"
output:
  html_document:
    df_print: paged
fontsize: 12pt
---

## (Note: This is spline version, need to change the get_fat functions)

## Goals

These are the in-class exercises we'll be doing for our special
emphasis on fertility (Wednesday). 

In the exercises, we'll focus on US fertility since the Great
Recession, 2008-2023. 

For your group project, you'll pick a different country from the Human
Fertility Database.

We're going to do 3 exercises:

1. Period-cohort translation (a la Ryder)

2. Period shift modeling (a la Bongaarts and Feeney)

3. Cohort shift modeling (a la Goldstein and Cassidy)

## 0. Preliminaries

Read in some packages and custom functions

```{r}
library(data.table)
source("utility_functions.R")
source("tempo_functions.R")
```

Read in data

```{r}
dt <- fread("asfrRR.txt")
dt <- dt[Code == "USA"] ## restrict to USA
```

Let's convert to matrix format f(a,t): fertility rates by 
age (rows) and time (columns)

```{r}
fat_all <- dt[, xtabs(ASFR ~ Age + Year)] ## create the matrix
tfr <- colSums(fat_all)
plot(names(tfr), tfr, ylim = c(1, 4), xlab = "Period", ylab = "TFR")
abline(v = 2008, lty = 2)
grid()
title("USA Period Total Fertility, since 1933")
```

You can see that since Great Recession, period fertility has been steadily
declining -- from about 2.2 in 2008 to about 1.6 in 2023.

For convenience we're going to limit ourselves to the period since
2008 and to ages 16-42. (Not much fertility occurs outside these ages,
so we're not missing much of the story.)

```{r}
## recreate smaller matrix
fat <- dt[Year >= 2008 & Age %in% 16:42, xtabs(ASFR ~ Age + Year)] 
tfr <- colSums(fat)
year <- as.numeric(colnames(fat))
plot(year, tfr, lty = 1, type = 'o', ylim = c(1.5,2.5), 
     xlab = "Period", ylab = "TFR")
grid()
title("USA Period Total Fertility, since 2008")
```

A big decline. 

* What happened about 2010 besides Great Recession? Hint: Obamacare.
* What happened from 2015 to 2019? 
* What happened in 2020?
* Has fertility stabilized, or would you expect continued decline?


### Explore the age-specific data

Now let's look at the changing age-specific rates

#### Changing age-specific rates

Time trend by age

```{r}
par(mfrow = c(1,1))
my_ages = paste(seq(18, 42, 6))
matplot(colnames(fat), t(fat[my_ages,]), type = 'l',
        ylab = "ASFR",
        xlab = "Period",
        main = "Age-specific fertility over time")
text(colnames(fat)[10], fat[my_ages,10], my_ages, col = 1:5)
```

* For which ages is fertility increasing?
* For which ages is fertility decreasing?
* Is this what you would expect if the whole age specific fertility 
  curve were shifting to older ages?

Age pattern by time

```{r}
my_periods = paste(seq(2008, 2023, 7))
matplot(rownames(fat), fat[,my_periods], type = 'l',
        lty = 1:3,
        col = "black",
        ylab = "ASFR",
        xlab = "Age",
        main = "Age-specific fertility schedules over time")
text(rownames(fat)[10], fat[10, my_periods], my_periods)
```

* What do you think about the age-pattern of fertility change?
* Does this look like only an age shift to the right or do you think 
  there's also been a level change?

### Changing mean ages?

```{r}
mt <- apply(fat, 2, get.mean)
plot(names(mt), mt, type = 'o',
     ylim = c(25, 30), xlab = "Period", ylab = "Mean age")
title("Period mean ages over time")
grid()
```

* Do you think it's the mean age itself or the rate of change in the mean age 
  that should affect the level of period fertility?

```{r}
mt.prime <- rt <- center.diff(mt, end.fill = TRUE)
plot(names(mt.prime), mt.prime, type = "p",
     ylim = c(0,.3), xlab = "Period", ylab = "Change in mean age")
title("Time derivative mean age")
grid()
```

We can see that mean age of birth rose from about 28 to about 30 over
the course of these 15 years, an average rate of change of 0.13 per
year.

* Is rate of postponement speeding up, slowing down, or staying about
  the same?
* Do you have some preliminary hypotheses about the role that timing has on 
  the observed decline in period TFR?

## 1. Period-cohort translation

Translation from periods to cohorts

We'll use the Ryder-inspired translation equation that goes from
observable period quantities to cohort total fertility 
(given on page 19 of the reading)

$$
CTFR(t - \mu_p) \approx TFR(t) \times (1 + \mu'_p(t))
$$

Estimate cohort TFR using the Ryder-inspired formula

```{r}
mu_p = mt ## period mean
mu_p_prime = rt ## time derivative of period mean
ctfr_hat = tfr * (1 + mu_p_prime) ## ryder-esque formula

## corresponding cohorts
t <- as.numeric(colnames(fat))
cohorts <- t - mu_p
names(ctfr_hat) <- cohorts
plot(names(ctfr_hat), ctfr_hat, xlab = "Cohort", 
     ylab = "Estimated CTFR", type = 'o',
     ylim = c(1.5, 2.5)) ## note: same ylim as with period before
grid()
```

* Cohort TFR looks like it's declining too. 

### Comparison of period and cohort

We can divide the Ryder-esque cohort estimates by the period TFR to
see how much higher the cohort TFR might be. (Note: We're ignoring the 
date shift here from period to cohort, isolating the $1 + \mu'(t)$ effect)

```{r}
coh_to_per_ratio <- ctfr_hat/tfr
names(coh_to_per_ratio) <- names(tfr)
print(round(coh_to_per_ratio,2))
```
Cohort TFR is higher by 5 to 17 percent.


### Comparison with extrapolated rates

Let's compare the Ryder-inspired estimate to "observed" cohort TFR. In order 
to get recent incomplete cohorts (e.g., those that are born after 1980), we'll
extrapolate recent age-specific rates forward in time (by about 10
years).

* If the oldest age is 42 and we observe fertility through 2023, what would 
  be the most recent cohort with completed fertility? 
  (This is the motivation for doing the extrapolation)

```{r}
## Extrapolation
N <- 10 ## number of years we are extrapolating for 
n_periods = ncol(fat)
bx <- ( fat[,n_periods] - fat[,n_periods - 5] ) / 5 ## rate of change calculated from most recent five years
fat_forward = fat[,n_periods] + cbind(bx) %*% rbind(1:N) ## extrapolate based on this rate of change staying constant in the future for N years
colnames(fat_forward) <- paste(max(as.numeric(colnames(fat))) + 1:N)
## now combine observed (form distant past) and extrapolation
fat_extended = cbind(fat_all[rownames(fat),], fat_forward)
```

Plot the extrapolated period rates.

```{r}
matplot(colnames(fat), t(fat), type='l', xlim = c(2008, 2035), xlab = "Period", 
        ylab = "ASFR", lty = 1)
matlines(colnames(fat_extended), t(fat_extended), lty = 2)
title("Linear Extrapolation of Period Age-Specific Fertility Rates")
```

* Looks pretty reasonable, right?

Now calculate cohort TFR

```{r}
## transform to a cohort matrix
fac_extended <- per2coh(fat_extended)
fac_extended <- remove.na.columns(fac_extended)

## compute CTFR of observed and extrapolated rates
ctfr_obs = colSums(fac_extended)

## compare "observed" and "translated"
plot(names(ctfr_obs), ctfr_obs, lty = 1, type = 'l',
     ylim = c(1, 3), lwd = 2,
     xlim = c(1970, 2000),
     xlab = "Cohort",
     ylab = "CTFR",
     main = "Cohort total fertility")
points(cohorts, ctfr_hat, lty = 2, cex = .6)
grid()

## we can also compare this to the "uninflated" period tfr
points(cohorts, colSums(fat), cex = .6, col = 'red')

legend("topright", c("obs CTFR",
                     "translated CTFR",
                     "PTFR"),
       lwd = c(2, -1, -1),
       cex = .6,
       pch = c(-1,1,1),
       col = c("black", "black", "red"),
       bty = "n")

```

How is the "translation" approach doing?
* fine "on average" (for about 1986)
* a bit too high at the beginning (when period fertility from 2008 is 
  being translated) 
* a bit too low at the end (when period fertility from 2023 is 
  being translated)

What else do we notice?
* cohort is about .2 children higher than period TFR (about 10%)
* the decline in period fertility is accompanied by a similar decline 
  in cohort fertility

Note: For recent cohorts our "observed" CTFR has quite a bit of
extrapolation.

Optional question (if we have time): 

* What would be some ways to get a *more robust* estimate of cohort
total fertility from recent period rates? (Hint: which quantities are
most unstable from one period to the next?)


## 2. Period shift modeling

### Tempo-adjusted period fertility

Our Bongaarts-Feeney formula for "tempo-adjusted TFR" is
$$
TFR^*(t) = {TFR(t) \over 1 - \mu'_p(t)}
$$

This is easy to implement, since we've already calculated all of the
ingredients.

```{r}
tfr_adj = tfr / (1 - rt)
```

We can plot this and see how it compares to the observed TFR.

```{r}
t = names(tfr)
plot(t, tfr, type = 'l', ylim = c(1.5, 2.5))
lines(t, tfr_adj, lty = 2)
legend("topright", c("tfr", "tfr_adj"),
       lty = 1:2)
grid()
```

The tempo-adjusted TFR has several features

a. Most importantly, it is larger than the observed TFR (for this time
      period in the USA).

b. It bounces about somewhat from year to year

c. Discuss if the following is TRUE:

	> The tempo-adjusted TFR* has declined at about the same rate as the
     observed TFR. This means that it doesn't appear that an
     acceleration of postponement is responsible for the recent
     declines in fertility

###  Consistency with change in age-specific schedules

The Bongaarts-Feeney model is a counter-factual measure, which has no
"truth" to which it can be compared. It tells us what observed period
fertility would have been in the absence of postponement, assuming
that postponement took the form of a uniform delay (by age) in
fertility from one period to the next.

We can assess the plausibility of such shifts, however, by looking at
what pattern of age-specific change would be consistent with the
Bongaarts-Feeney model. The Goldstein and Cassidy chapter tells us
that an age-period surface of fertility consistent with the
Bongaarts-Feeney model is given by:

$$
f(a,t) = f_0(a - R(t)) \times (1-R'(t)) \times q(t),
$$
where $R(t)$ is the total number of years by which women in year $t$ have shifted their fertility, $R'(t)$ is
the rate of shift in year $t$, and $q(t)$ is the period "quantum",
equal to the tempo-adjusted fertility $TFR^*(t)$. 

###  Numerical optimization 

We can estimate all of the parameters of this model using numerical
optimization. In R, one way to do this is the nlm() function. The way
optimization works is that the computer tries different values of the
parameters until it finds the "best" ones. In this case, the "best" is
defined by parameters  $q_t$, $f_0$, and $R_t$ that produce estimates
$\hat{f}(a,t)$ that are close to the observed fertility rates.

First we define the function that implements the period shift model.

```{r}
get_fat_bf <- function(qt, f0, Rt)
{
    ## period-shift estimated fertility surface f(a,t) from parameters
    ## -- shifts all ages of period at once
    ## -- indices without names

    ## get rt (time derivative of Rt)
    rt = center.diff(Rt, end.fill = T)

    ## initialize empty period matrix
    fat_hat <- matrix(NA, nrow(fat), ncol(fat))
    dimnames(fat_hat) <- dimnames(fat)
    a <- as.numeric(names(f0))

    ## new spline function
    f0_spline <- splinefun(x = a, y = f0, method = "natural")
    
    ## fill up empty matrix with shifted f0`
    for (i in 1:ncol(fat)) ## each period
    {
        ## rule 2, just extrapolates edge values
        ## a problem if ages start late or end early
        ## this_shifted_fat <-
        ##      approx(x = a, y = f0,
        ##             xout = a - Rt[i], rule = 2)$y
        shifted_a = a - Rt[i]
        this_shifted_fat <- f0_spline(shifted_a)

        fat_hat[,i] <- this_shifted_fat * (1 - rt[i])
    }
    ## multiply by qt (using diag trick)
    fat_hat <- fat_hat %*% diag(qt)

    ## return fat_hat
    return(fat_hat)
}
```

Then, we'll define the "objective" function that will tell us how close
our estimated fertility rates are to the observed ones:

```{r}
ob_fun_f0_qt_Rt <- function(p)
{
    ## objective function for cohort shift model
    
    ## get latest version of parameters
    n_a <- length(f0_start)
    n_t <- length(qt_start)
    f0 = exp(p[1:n_a])
    f0 <- f0/sum(f0) ## normalize
    qt = exp(p[(n_a + 1):(n_a + n_t)])
    Rt = p[(n_a + n_t + 1):length(p)]
    names(Rt) <- names(Rt_start)
    names(f0) <- names(f0_start)
    names(qt) <- names(qt_start)

    ## get the f(a,t) implied by these parameters
    fat_hat = get_fat_bf(qt, f0, Rt) ## implied period shift f(a,t)
    dimnames(fat_hat) <- dimnames(fat)

    ## define loss  and return it (to optimizer)
    loss = sum((fat-fat_hat)^2) ## sum of squared errors
    return(loss)
}
```

Now we can use the optimizer. This involves starting with some
reasonable guesses at the parameter values and running nlm().

```{r}
## starting values
## -- f0
f0_start <- rowMeans(fat)
f0_start = f0_start/sum(f0_start) ## normalized
## -- qt
tfr <- colSums(fat)
qt_start <- tfr
## -- Rt
mu_t <- apply(fat, 2, get.mean)
Rt_start <- mu_t - mean(mu_t)


## optimizer (this takes a while!)
p_start = c(log(f0_start), log(qt_start), Rt_start)
fit = nlm(p = p_start,
          f = ob_fun_f0_qt_Rt, print.level = 1,
          iterlim = 1000)
fit_bf = fit
```

Plot parameter estimates. The first section of the plot (between the first two 
dashed lines) shows optimized values for $log(f_0)$, the second section shows 
estimates for $log(q_t)$, and the third shows estimates for $log(R_t)$. 
The log scale is used here so that we ensure values for $f_0$, $q_t$, and $R_t$ 
are greater than 0. 

```{r}
## plot estimates
plot(fit$est,
       xlab = "Index in Parameter Vector",
       ylab = "Parameter Value")
abline(v = cumsum(c(1, length(f0_start),
                    length(qt_start),
                    length(Rt_start))),
       lty = 2
       )
```

Compare fitted to observed $f(a,t)$

```{r} 
fit = fit_bf
f0_hat = exp(fit$est[1:length(f0_start)])
qt_hat = exp(fit$est[(length(f0_start) + 1):(length(f0_start) + length(qt_start))])
qt_hat_bf = qt_hat
Rt_hat = fit$est[(length(f0_start) + length(qt_start) + 1):length(fit$est)]
## normalize
f0_hat = f0_hat/sum(f0_hat)


names(f0_hat) <- names(f0_start)
names(qt_hat) <- names(qt_start)
names(Rt_hat) <- names(Rt_start)


fat_hat = get_fat_bf(qt = qt_hat, f0 = f0_hat, Rt = Rt_hat)
colnames(fat_hat) <- colnames(fat)

a <- rownames(fat)
par(mfrow = c(2,2))
plot(a, fat[,"2010"], cex = .6, pch = 19)
lines(a, fat_hat[,"2010"], lty = 2)
title("2010")
##
plot(a, fat[,"2016"], col = 2, , cex = .6, pch = 19)
lines(a, fat_hat[,"2016"], lty = 2, col = 2)
title("2016")
##
plot(a, fat[,"2022"], col = 3, cex = .6, pch = 19)
lines(a, fat_hat[,"2022"], lty = 2, col = 3)
title("2022")
## 
my_periods = paste(c("2010", "2016", "2022"))
matplot(a, fat[, my_periods], lty = 1, type = 'p', pch = 10, cex = .6)
matlines(a, fat_hat[, my_periods], lty = 2)
title("All three periods")
```

Looks like a very good fit.


Check against Bongaarts-Feeney tempo-adjusted TFR

```{r}
par(mfrow = c(1,1))
plot(t, tfr, ylim = c(1, 3), xlab = "Year", ylab = "TFR")
lines(t, tfr_adj, lty = 2)
lines(t, qt_hat_bf, lty = 2, col = 'red')
legend("topright", 
       c("observed tfr", "adj tfr (bongaarts-feeney)", 
         "adj tfr (optimization)"),
       col = c("black", "black", "red"),
       lty = c(-1, 2, 2),
       pch = c(1,-1,-1),)
```

Good match?

## 3. Cohort shift modeling

###  Estimation of cohort shift model

The cohort shift model is given by
$$
f(a,t) = f_0(a - S(t-a)) \times q(t),
$$
where $f_0$ is a baseline fertility schedule, normalized to add up to
1.0, $a - S(t-a)$ is the shifted age for each cohort $t-a$, and $q(t)$
is the quantum of fertility in the absence of shifts, in this case
Goldstein and Cassidy's $TFR^\dagger$.

Once again, we can estimate all of the parameters of the model using numerical
optimization. 

First we define the function that implements our cohort shift model

```{r}
get_fat_gc <- function(qt, f0, Sc)
{
    ## cohort-shift estimated fertility surface f(a,t) from parameters
    ## -- shifts all ages of cohort at once
    ## -- indices without names
    ## -- converts back to period

    ## initialize empty cohort matrix
    fac_hat <- matrix(NA, nrow(fac), ncol(fac))
    dimnames(fac_hat) <- dimnames(fac)
    a <- as.numeric(names(f0))

    ## new spline function
    f0_spline <- splinefun(x = a, y = f0, method = "natural")

    
    ## fill up empty matrix with shifted f0
    for (i in 1:ncol(fac)) ## each cohort
    {
        ## old:
        ## rule 2, just extrapolates edge values
        ## a problem if ages start late or end early
        ## this_shifted_fac <-
        ##      approx(x = a, y = f0,
        ##             xout = a - Sc[i], rule = 2)$y

        ## new: 
        shifted_a = a - Sc[i]
        this_shifted_fac <- f0_spline(shifted_a)

        fac_hat[,i] <- this_shifted_fac
    }
    ## convert to period
    fat_hat <- coh2per(fac_hat)
    fat_hat <- fat_hat[, colnames(fat)] ## keep period range
    
    ## multiply by qt (using diag trick)
    fat_hat <- fat_hat %*% diag(qt)

    ## return fat_hat
    return(fat_hat)
}
```

Then, we'll define the "objective" function that will tell us how close
our estimated fertility rates are to the observed ones:

```{r}
ob_fun_f0_qt_Sc <- function(p)
{
    ## objective function for cohort shift model
    
    ## get latest version of parameters
    n_a <- length(f0_start)
    n_t <- length(qt_start)
    f0 = exp(p[1:n_a])
    f0 <- f0/sum(f0) ## normalize
    qt = exp(p[(n_a + 1):(n_a + n_t)])
    Sc = p[(n_a + n_t + 1):length(p)]
    names(Sc) <- names(Sc_start)
    names(f0) <- names(f0_start)
    names(qt) <- names(qt_start)

    ## get the f(a,t) implied by these parameters
    fat_hat = get_fat_gc(qt, f0, Sc) ## implied cohort shift f(a,t)
    dimnames(fat_hat) <- dimnames(fat)

    ## define loss  and return it (to optimizer)
    loss = sum((fat-fat_hat)^2) ## sum of squared errors 
    return(loss)
}
```

Again, we'll use the optimizer, nlm(). 

First, we'll introduce a faster utility function to speed things
up. Don't worry about how this works :)

```{r}
## faster coh2per
coh2per_fast_minimal <- function(Acoh) {
  a_vals <- as.integer(rownames(Acoh))
  c_vals <- as.integer(colnames(Acoh))

  n_age <- length(a_vals)
  n_coh <- length(c_vals)

  t_vals <- outer(a_vals, c_vals, `+`)
  f_vals <- Acoh

  t_range <- range(t_vals)
  t_seq <- t_range[1]:t_range[2]

  Aper <- matrix(NA_real_, nrow = n_age, ncol = length(t_seq),
                 dimnames = list(as.character(a_vals), as.character(t_seq)))

  Aper[cbind(rep(seq_len(n_age), n_coh), as.integer(t_vals - t_seq[1] + 1))] <- f_vals
  ## +1 is because indexing doesn't start at 0 in R

  Aper
}
coh2per <- coh2per_fast_minimal
```



```{r}
fac <- per2coh(fat)
fac <- remove.na.columns(fac)
## starting values
## -- f0
f0_start <- rowMeans(fat)
f0_start = f0_start/sum(f0_start) ## normalized
## -- qt
tfr <- colSums(fat)
qt_start <- tfr
## -- Sc from Rt
mu_t <- apply(fat, 2, get.mean)
Rt <- mu_t - mean(mu_t)
#plot(Rt)
Sc_start <- approx(x = names(Rt),
                   y = Rt,
                   xout = as.numeric(colnames(fac)) + 30,
                   rule = 2)$y
names(Sc_start) <- colnames(fac)

## optimizer (this takes a while!)
p_start = c(log(f0_start), log(qt_start), Sc_start)
if(!exists("gc_fit"))
{
    fit = nlm(p = p_start,
              f = ob_fun_f0_qt_Sc, print.level = 1,
              iterlim = 1000)
    gc_fit = fit
}
print(gc_fit)
```

And we'll plot the parameter estimates, $log(f_0)$, $log(q_t)$, and $S_c$. 

```{r}
plot(fit$est)
abline(v = cumsum(c(1, length(f0_start),
                    length(qt_start),
                    length(Sc_start))),
       lty = 2)
```

A cleaner visualization of each parameter separately
```{r}
p_hat = gc_fit$estimate
n_a = length(f0_start)
n_t = length(qt_start)
n_c = length(Sc_start)
year = names(qt_start)
age = names(f0_start)
cohort = names(Sc_start)
f0_hat = exp(p_hat[1:n_a])
## normalize f0_hat!!! 
f0_hat = f0_hat/sum(f0_hat)
qt_hat = exp(p_hat[(n_a + 1):(n_a + n_t)])
Sc_hat = p_hat[(n_a + n_t + 1):length(p_hat)]
## apply names
names(Sc_hat) <- names(Sc_start)
names(f0_hat) <- names(f0_start)
names(qt_hat) <- names(qt_start)
plot(age, f0_hat, type = 'o', col = 'red',
     main = "Estimated cohort age-density f_0(a)",
     ylab = "Proportion")
grid()
```

```{r}
plot(cohort, Sc_hat, col = 'red',
     main = "Estimated cohort shifts S(c)",
     ylab = "Cumulative years of postponement")
grid()
```

```{r}
plot(year, qt_hat, type = 'o', col = 'red', ylim = c(1, 3),
     main = "Estimated period quantum vs. TFR",
     ylab = "Children per woman")
lines(year, tfr)
grid()
```




### Compare fitted to observed $f(a,t)$

```{r} 
fat_hat = get_fat_gc(qt = qt_hat, f0 = f0_hat, Sc = Sc_hat)
dimnames(fat_hat) <- dimnames(fat)
sum((fat_hat - fat)^2)
## [1] 0.001652185
```

Plot results 

```{r}
a <- rownames(fat)
par(mfrow = c(2,2))
plot(a, fat[,"2010"], cex = .6, pch = 19)
lines(a, fat_hat[,"2010"], lty = 2)
title("2010")
##
plot(a, fat[,"2016"], col = 2, , cex = .6, pch = 19)
lines(a, fat_hat[,"2016"], lty = 2, col = 2)
title("2016")
##
plot(a, fat[,"2022"], col = 3, cex = .6, pch = 19)
lines(a, fat_hat[,"2022"], lty = 2, col = 3)
title("2022")
## 
my_periods = paste(c("2010", "2016", "2022"))
matplot(a, fat[, my_periods], lty = 1, type = 'p', pch = 10, cex = .6)
matlines(a, fat_hat[, my_periods], lty = 2)
title("All three periods")
```

Ok, this also looks very good.

### Comparison of observed TFR with period- and cohort-tempo-adjusted measures


```{r}
par(mfrow = c(1,1))
plot(t, tfr, type = 'p', ylim = c(1.5, 2.5))
lines(t, qt_hat_gc, col = 'blue', lty=2)
lines(t, qt_hat_bf, col = 'red', lty=2)
legend("topright",
       c("tfr", "period-tempo-adjusted tfr",
         "cohort-tempo-adjusted tfr"),
       col = c("black", "red", "blue"),
       lty = c(-1, 2, 2),
       pch = c(1, -1, -1))
grid()
```


Questions: 

1. (important) Does either the period-shift or cohort-shift model give any reason to
   believe that postponement (or perhaps changes in the rate of
   postponement) is behind the recent drop in fertility?
   
2. (important) Is the recent drop in fertility below age 25 consistent with period
   shifts and/or, cohort shifts?
   
3. (optional) What would the cohort TFR look like if we plotted it at the time
   when the cohort reaches its mean age of childbearing? (Hint: very much
   like one of these lines -- which one?) 
   
4. (optional) What might we be missing by looking at fertility of all parities at
   once rather than doing a parity specific analysis?
   
5. (optional) What about the cohort model makes it more difficult to estimate?

a. the incompleteness of cohorts means we can't just estimate changes
in the mean in a simple way
b. period changes in fertility levels q(t) influence cohort age
schedules. So, we need to estimate q and f0 at the same time.
c. both a & b
d. something else entirely?


## 4. Group exercises/project

Here's a plot of TFR in a number of HFD countries. Pick one of these
countries to analyze. 

```{r}
dt <- fread("./asfrRR.txt")

tfr.dt <- dt[Year %in% 2008:2023, .(tfr = sum(ASFR)), by = .(Code, Year)]

my_codes <- c("CAN", 
    "ESP",
    "FIN", "JPN", "KOR",
    "NOR", "SWE", "USA") 
tfr.mat <- tfr.dt[Code %in% my_codes, 
                  xtabs(tfr ~ Code + Year)]
tfr.mat[tfr.mat == 0] <- NA
matplot(2008:2023, t(tfr.mat), type = 'l', lty = c(1),
        col = 1:7, xlim = c(2007, 2023),
        xlab = "Year",
        ylab = "Period TFR",
        main = "Recent TFRs of selected HFD countries")
grid()
text(x = 2008, y = tfr.mat[,"2008"],
     pos = 2,
     rownames(tfr.mat),
     col = 1:7)
```

(a) Repeat some or all of the analysis we did in the USA for your country,
applying Ryder, Bongaarts-Feeney, and Goldstein-Cassidy approaches. Do changes
in timing appear to be behind recent fertility decline? Pick your most 
informative graphic to share.

(b) (optional) Try something else of your own invention with the
observed age-specific rates f(a,t) for your country in order to shed
light on the nature of the how fertility is declining in that country.

(c) (required) Discuss whether you think the following statement is
true for your country: 

> "The fact that fertility decline is occuring mostly at younger ages,
with little to no offsetting increase at older ages, means that fertility 
is being foregone, not just delayed."



